---
title: "AnnData interop"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AnnData interop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(dafr)
```

# DafAnnData facade

If you've been using `anndata` / `scanpy` / `Seurat`, the facade
exposes a Daf with familiar property names:

```{r}
d <- example_cells_daf()
ann <- as_anndata(d)
ann$n_obs
ann$n_vars
head(ann$obs_names)
dim(ann$X)
```

The facade is read-only: `ann$X <- ...` errors. To modify data, write
to the underlying Daf: `set_matrix(d, ...)`.

# Loading h5ad

```{r, eval = FALSE}
d <- h5ad_as_daf("path/to/file.h5ad")
```

Requires `hdf5r` (a Suggests dep): `install.packages("hdf5r")`.

# Writing h5ad

```{r, eval = FALSE}
daf_as_h5ad(d, "out.h5ad", overwrite = TRUE)
```

# Fixture round-trip

```{r}
fixture <- system.file("extdata", "small_test.h5ad", package = "dafr")
if (file.exists(fixture) && requireNamespace("hdf5r", quietly = TRUE)) {
    d <- h5ad_as_daf(fixture)
    cat("loaded:", daf_name(d), "\n")
    cat("n_obs:", length(axis_entries(d, "obs")), "\n")
    cat("n_vars:", length(axis_entries(d, "var")), "\n")
}
```

# What the h5ad readers translate

`h5ad_as_daf()` (and `daf_as_h5ad()` in the other direction) cover:

- `/X` dense and sparse (`csr_matrix`, `csc_matrix`) - sparse is
  round-tripped via `Matrix::sparseMatrix`.
- `/obs` and `/var` columns: plain datasets and `categorical` groups
  (read as `factor`, ordered or not).
- `/layers` - dense layers become extra `obs x var` matrices keyed by
  the layer name.
- `/obsm` and `/varm` dense matrices - each is placed on a synthetic
  axis named `obsm_<name>_dim` / `varm_<name>_dim`.
- `/uns` scalars (flat and nested; nested keys are flattened with
  `_`).

Things still NOT translated (escalated via `unsupported_handler`,
default `WARN_HANDLER`, then skipped):

- `/obsp` and `/varp` (square per-pair matrices on obs/var).
- `/raw` (the legacy raw counts slot).
- Sparse layers / `obsm` / `varm` entries; only dense forms are read.
- Group-valued obs/var columns other than `categorical`.

If a strict round-trip matters, pass `unsupported_handler =
ERROR_HANDLER` so unhandled entries hard-fail instead of being
dropped silently.
