---
title: "Getting Started with dafr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with dafr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(dafr)
```

## What is a Daf?

A **Daf** (Data Axes Format) is a typed, multi-axis container for
scientific data. Each daf holds:

- **Axes** - named lists of entries (e.g. `cell`, `gene`, `donor`).
- **Scalars** - single values (e.g. `organism = "human"`).
- **Vectors** - one value per entry on a chosen axis.
- **Matrices** - one value per pair of entries on two chosen axes.

Think of it as AnnData generalised to any number of axes, with
typed scalars, on-disk persistence in several formats, and a query
DSL for composing reads.

## Build a daf in memory

```{r}
d <- memory_daf(name = "demo")
add_axis(d, "cell", c("c1", "c2", "c3"))
add_axis(d, "gene", c("g1", "g2"))
set_scalar(d, "organism", "human")
set_vector(d, "cell", "donor", c("A", "B", "A"))
set_matrix(d, "cell", "gene", "UMIs",
           matrix(1:6, nrow = 3, ncol = 2))
print(d)
```

`description(d)` produces a more detailed dump of axes, vectors,
matrices, and scalars, useful for inspecting unfamiliar dafs:

```{r}
cat(description(d))
```

## Read data back

```{r}
get_scalar(d, "organism")
get_vector(d, "cell", "donor")
get_matrix(d, "cell", "gene", "UMIs")
```

`scalars_set(d)`, `vectors_set(d, axis)`, and `matrices_set(d, a, b)`
list the property names present on the corresponding container.

## Queries

Queries let you compose reads. Two equivalent forms - a string DSL
and pipe-chain builders - share the same execution engine:

```{r}
d[". organism"]
d[Axis("cell") |> LookupVector("donor")]
```

```{r}
# Mean UMI count per cell (reduce across the gene axis):
d["@ cell @ gene :: UMIs >| Mean"]
```

See `vignette("queries", package = "dafr")` for the practical tour
and `vignette("query-dsl-reference", package = "dafr")` for the full
operator grammar.

## Data frames and dplyr

`get_dataframe()` materialises an axis-keyed slice as a `tibble`:

```{r}
get_dataframe(d, "cell")
```

The dplyr backend lets you treat any axis as a `tbl` and pipe verbs:

```{r, eval = requireNamespace("dplyr", quietly = TRUE)}
library(dplyr, warn.conflicts = FALSE)
tbl(d, "cell") |>
    group_by(donor) |>
    summarise(n = n())
```

See `vignette("dplyr", package = "dafr")` for the supported verbs and
write-back semantics.

## Persistence

A daf can be persisted in three on-disk shapes - pick the one that
matches how the data needs to travel.

```{r}
# files_daf: one file per property, mmap-backed reads.
path <- tempfile("dafr-")
fd <- files_daf(path, mode = "w+", name = "persisted")
copy_all(fd, d)  # destination first, then source
list.files(path, recursive = TRUE)
```

```{r}
# Reopen read-only:
fd2 <- files_daf(path, mode = "r")
get_scalar(fd2, "organism")
get_matrix(fd2, "cell", "gene", "UMIs")
```

`zarr_daf()` (directory or `.daf.zarr.zip`) and `http_daf()`
(read-only HTTP) round-trip the same data; see
`vignette("zarr", package = "dafr")`.

## Where to go next

- **Query DSL** - `vignette("queries", package = "dafr")`.
- **dplyr backend** - `vignette("dplyr", package = "dafr")`.
- **AnnData / h5ad interop** - `vignette("anndata", package = "dafr")`.
- **Zarr & HTTP storage** - `vignette("zarr", package = "dafr")`.
- **Native performance notes** - `vignette("native-performance", package = "dafr")`.
- **Larger building blocks** - chains (`chain_reader`), views
  (`viewer`, `VIEW_ALL_DATA`), contracts (`Contract`, `verify_input`),
  and computations (`computation`, `adapter`) all have detailed help
  pages in the reference.

## Example data

```{r}
d2 <- example_cells_daf()
axes_set(d2)
```
