---
title: "Getting Started with dafr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with dafr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(dafr)
```

## What is a Daf?

A Daf (Data Axes Format) is a multi-axis container for biological /
scientific data. Think of it as a typed dictionary of scalars,
per-axis vectors, and per-axis-pair matrices, with cache invalidation
and lazy mmap-backed reads.

## Create a Daf

```{r}
d <- memory_daf(name = "demo")
add_axis(d, "cell", c("c1", "c2", "c3"))
add_axis(d, "gene", c("g1", "g2"))
set_scalar(d, "organism", "human")
set_vector(d, "cell", "donor", c("A", "B", "A"))
set_matrix(d, "cell", "gene", "UMIs",
           matrix(1:6, nrow = 3, ncol = 2))
print(d)
```

## Reading data

```{r}
get_scalar(d, "organism")
get_vector(d, "cell", "donor")
get_matrix(d, "cell", "gene", "UMIs")
```

## Queries

Queries let you compose reads. Two equivalent forms — a string DSL and
pipe-chain builders:

```{r}
get_query(d, "@ cell : donor")
get_query(d, ". organism")
```

```{r}
d[Axis("cell") |> LookupVector("donor")]
```

## Data frames

```{r}
get_dataframe(d, "cell")
```

## Persistence

```{r, eval = FALSE}
# Write to a directory:
fd <- files_daf(tempfile("dafr-"), mode = "w+", name = "persisted")
copy_all(d, fd)
```

## Example data

```{r}
d2 <- example_cells_daf()
axes_set(d2)
```
