---
title: "dplyr backend"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dplyr backend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(dafr)
```

dafr exposes a [dplyr](https://dplyr.tidyverse.org) backend for the
**axes** of a daf. Each axis becomes a `tbl` whose rows are the axis
entries and whose columns are the vectors defined on that axis.
Matrices are intentionally *not* exposed through this interface — use
the native query DSL (`vignette("queries", package = "dafr")`) for
those.

```{r, eval = requireNamespace("dplyr", quietly = TRUE)}
library(dplyr)
d <- example_cells_daf()
tbl(d, "cell")
```

## Familiar verbs

`filter`, `select`, `mutate`, `arrange`, `summarise`, `group_by`,
`distinct`, and `pull` all work as you'd expect. `collect()`
materializes the tbl as a tibble.

```{r, eval = requireNamespace("dplyr", quietly = TRUE)}
tbl(d, "cell") |>
    filter(!is.na(donor)) |>
    group_by(donor) |>
    summarise(n = n()) |>
    arrange(desc(n))
```

## Write-back is explicit

`mutate()` stores computed columns in memory; it does not touch the
daf. Persisting them as vectors on the axis requires an explicit
`compute(vectors = ...)` — dafr never silently writes to the daf.

```{r, eval = FALSE}
tbl(d, "cell") |>
    mutate(log_umis = log10(n_umis)) |>
    compute(vectors = "log_umis")
```

The daf backend reuses `dplyr::compute()` (with an extra
`vectors = ...` argument) rather than introducing a new generic, so
it doesn't shadow `dplyr::compute()` for dbplyr users.

`compute()` errors if the tbl has been filtered (partial row mask) —
you cannot write back a partial vector. A permuted-but-full row mask
(e.g. after `arrange()`) is un-permuted to axis order before writing.

## Grouping that looks up an axis

If the grouping variable names an existing axis in the daf, the
result of `summarise()` is a `daf_axis_tbl` keyed to that axis. You
can keep piping dplyr verbs, and later `compute()` the derived
columns back as vectors on that axis.

```{r, eval = FALSE}
# "donor" is an axis; the summarise result is itself tbl(d, "donor")-like.
tbl(d, "cell") |>
    group_by(donor) |>
    summarise(mean_umis = mean(n_umis)) |>
    compute(vectors = "mean_umis")  # persists on the donor axis
```

If the grouping variable is just a vector (not an axis), `summarise()`
returns a plain tibble.

## Out of scope for v1

- Matrices. Use `Axis(...) |> LookupMatrix(...)` from the query DSL.
- Joins across axes.
- Window functions (`lag`, `cumsum`, etc.).
- DSL pushdown for `filter` / `summarise` (planned follow-up — verbs
  currently materialize via `get_vector`, which is cheap for memory
  daf but not optimal for large `FilesDaf` with selective filters).
