---
title: "Query DSL"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Query DSL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(dafr)
d <- example_cells_daf()
```

dafr ships a small query language for composing reads against a daf.
This vignette is the practical tour; for the full operator grammar see
`vignette("query-dsl-reference", package = "dafr")`.

# Two equivalent forms

Every query has two surface syntaxes that compile to the same plan:

- **String DSL** - terse, good for ad-hoc work and scripts.
- **Builder pipe-chains** - composable R objects; good when query
  fragments are computed at runtime.

```{r}
# String:
head(d["@ cell : donor"])
# Builder:
head(d[Axis("cell") |> LookupVector("donor")])
```

`canonical_query()` and `parse_query()` let you inspect what a string
compiles to:

```{r}
canonical_query("@ cell : donor")
```

# Lookups

The smallest queries look up a single piece of data. Each is keyed by
how many axes you push onto the stack before the property.

```{r}
# Scalar (zero axes):
d[". organism"]
# Vector (one axis):
head(d["@ donor : age"])
# Matrix (two axes):
dim(d["@ cell @ gene :: UMIs"])
```

# Masks: filtering an axis

Wrap a comparison in `[ ... ]` (string) or `BeginMask(...)` /
`EndMask()` (builder) to keep only the entries that pass:

```{r}
# Donors older than 60 - string and builder are equivalent:
length(d["@ donor [ age > 60 ]"])
length(d[Axis("donor") |> BeginMask("age") |> IsGreater(60) |> EndMask()])
```

Masks compose with `&` (AND), `|` (OR), `^` (XOR), each with an
optional `!` for negation. Combinators are left-to-right, not by
precedence - parenthesize via nested `[ ... ]` blocks if you need
grouping.

```{r}
# Female donors over 60:
length(d["@ donor [ age > 60 & sex = female ]"])
length(d[Axis("donor") |>
    BeginMask("age") |> IsGreater(60) |>
    AndMask("sex") |> IsEqual("female") |>
    EndMask()])
```

# Element-wise transforms

`Log`, `Abs`, `Clamp`, `Round`, `Convert`, etc. transform the values
in place. In the string DSL they follow a `%` prefix.

```{r}
# log2(age + 1):
head(d["@ donor : age % Log base 2 eps 1"])
head(d[Axis("donor") |> LookupVector("age") |> Log(base = 2, eps = 1)])
```

# Reductions

A reduction collapses one dimension. For matrices, `>-` reduces
**to a row** (across the column axis), `>|` reduces **to a column**
(across the row axis); the builder forms are `ReduceToRow()` and
`ReduceToColumn()`. For vectors the reduction is `>>` and yields a
scalar.

```{r}
# Per-gene mean UMIs (reduce across cells):
head(d["@ gene @ cell :: UMIs >- Mean"])
head(d[Axis("gene") |> Axis("cell") |> LookupMatrix("UMIs") |> ReduceToRow(Mean())])
```

```{r}
# Library size per cell (reduce across genes):
head(d["@ cell @ gene :: UMIs >| Sum"])
```

```{r}
# Vector -> scalar:
d["@ donor : age >> Mean"]
```

# GroupBy

`/` on a vector groups by another property of the same length and
applies a reduction per group. `-/` / `|/` do the same for matrix
rows / columns.

```{r}
# Mean donor age, split by sex:
d["@ donor : age / sex >> Mean"]
d[Axis("donor") |> LookupVector("age") |> GroupBy("sex", AsAxis()) |> Mean()]
```

```{r}
# Mean UMIs per donor, per gene (group cell rows by their donor):
m <- d["@ cell @ gene :: UMIs -/ donor >- Mean"]
dim(m)
m[1:3, 1:3]
```

# IfMissing: safe lookups

`IfMissing()` lets a vector / matrix lookup fall back to a default
when the property doesn't exist. Useful for queries that survive
across dafs with slightly different schemas.

```{r}
head(d[Axis("gene") |> LookupVector("absent_property") |> IfMissing(-1)])
```

# Tips

- Use `parse_query(s)` to see the operator list a string compiles to;
  this is the easiest way to learn the grammar by analogy.
- Builder fragments are ordinary R objects: store them in variables,
  build them in loops, pass them around. The expression
  `Axis("cell") |> LookupVector("donor")` is a `DafrQuery` you can
  index with later.
- `?Axis`, `?LookupVector`, `?Mean`, `?GroupBy` (etc.) have the full
  per-operator detail; the reference grouping is also in the pkgdown
  "Query builders" section.
